Generative AI for Image Reconstruction with Intensity Interferometry: A First Attempt

### 2024-10-02 Phase_II/draft/GAN.tex
second edit

GANs can be extended to a conditional model \citep{mirza2014conditional}. In this case, both generator and discriminator receive extra information $y$, such that the value function of the conditional GAN (cGAN) can be expressed as:

@@

Furthermore, \citep{isola2017image} observed that combining the cGAN given in equation \ref{eq:conditional_GAN} with the traditional L1-loss improves results, as the generator tries to be close to the ground truth. Hence, the function that is minimized as:
\begin{equation}
	\centering
	L_{tot} = \arg \min_{G} \max_{D} V(D, G) + \lambda \cdot L_1(G)
	\label{eq:total_gen_loss}
\end{equation}
where $\lambda$ = 100 and $L_1(g)$ is given by equation (\ref{eq:l1_loss})
\begin{equation}
	\centering
	L_1(g) = \mathbb{E}_{x,y,z} \left[ ||{y - G(x,z)}||_1 \right]
	\label{eq:l1_loss}
\end{equation}

This type of network is shown to be very robust, as it has been applied to several problems. Examples include creating colored images from grayscale images, creating images of facades based on architectural labels, changing from day to night in different pictures, and predicting a map based on satellite data. A longer list of applications is given in \citep{isola2017image}.

\subsection{Generator}
The used generator in this paper is a common U-Net convolutional network \citep{ronneberger2015u}. It consists of a contracting part, reducing the image size, and an expansive part, where the image is enlarged. The contracting part consists of repeated application of alternating convolutional layers and rectified linear unit (ReLU) layers. The expansion consists of repeated convolution, batch normalization, and ReLU layers. The first contracting step also includes a dropout layer \citep{isola2017image}

>>>>

GAN can be extended to a conditional model \citep{mirza2014conditional}. In this case, both generator and discriminator receive extra information $y$, such that the value function of the conditional GAN (cGAN) can be expressed as:

@@

Furthermore, \cite{isola2017image} observed that combining the cGAN given in equation \ref{eq:conditional_GAN} with the traditional L1-loss improves results, as the generator tries to be close to the ground truth. Hence, the function that is minimized as:
\begin{equation}
	\centering
	L_{tot} = \arg \min_{G} \max_{D} V(D, G) + \lambda \cdot L_1(G)
	\label{eq:total_gen_loss}
\end{equation}
where $\lambda$ = 100 and $L_1(G)$ is given as
\begin{equation}
	\centering
	L_1(G) = \mathbb{E}_{x,y,z} \left[ ||{y - G(x,z)}||_1 \right]
	\label{eq:l1_loss}
\end{equation}

This type of network is shown to be very robust, as it has been applied to several problems. Examples include creating colored images from grayscale images, creating images of facades based on architectural labels, changing from day to night in different pictures, and predicting a map based on satellite data. A longer list of applications is given in \cite{isola2017image}.

\subsection{Generator}
The used generator in this paper is a common U-Net convolutional network \citep{ronneberger2015u}. It consists of a contracting part, reducing the image size, and an expansive part, where the image is enlarged. The contracting part consists of repeated application of alternating convolutional layers and rectified linear unit (ReLU) layers. The expansion consists of repeated convolution, batch normalization, and ReLU layers. The first contracting step also includes a dropout layer \citep{isola2017image}.

### 2024-10-02 Phase_II/draft/II.tex
second edit

\section{Intensity Interferometry with ICTA}
Recently, Intensity Interferometry (II) has emerged as an additional tool to resolve stellar objects as it solves some of the key limitations of classical optical interferometry (atmospheric turbulence, length of baseline, and the sensitivity of the optical aperture). It is suitable for the high-resolution study of very massive, hot, and short-lived stars, such as OB-type, Wolf-Rayet, and Pulsating stars in N-body systems. R. Hanbury Brown and Robert Q. Twiss first installed the Narrabri Stellar Intensity Interferometer (NSII) in 1974 to resolve the diameter of bright stars at visible wavelengths \citep{brown1954lxxiv, brown1974intensity}. Nowadays, Cherenkov Telescope Arrays (CTAs), which are primarily used for gamma-ray observations, are at the forefront of implementing II during moonlit nights when it does not detect gamma-ray radiation. Intensity correlation and angular size measurements of stars have been detected, particularly using Very Energetic Radiation Imaging Telescope Array Systems (VERITAS), High Energy Stereoscopic Systems (HESS) and Major Atmospheric Gamma Imaging Cherenkov Telescopes (MAGIC) \citep{kieda2022performance, zmija2022optical, abe2024performance}.

@@

\subsection{The Signal-to-Noise Ratio for II}
\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig/telescope.png}
	\caption{The telescope configuration with similar properties each used to simulate the signal for II.}
	\label{fig:teles}
\end{figure}
The main physics channel of ICTAs is the study of Very High Energy gamma rays ($\geq 30$ GeV) originating from particle showers in the atmosphere. Telescopes have an array of mirrors, which focus the light onto an array of photo-multiplier tubes (PMTs) \citep{aleksic2016major}. Here, fig.~\ref{fig:teles} shows an arrangement of four IACTs with similar properties each. For II, the light of a stellar source is focused on a single PMT, which has a filter in front. The purpose of this filter is to efficiently transmit light around mean observational frequency, which protects the PMTs from excessive light. It is necessary, as II observations are mainly done during full moon nights, where it is too bright for Very High Energy $\gamma$-ray observations. The optical signal of the PMTs is transmitted with optical fibers, converted to an electrical signal, and digitized \citep{acciari2020optical}. The measurable observable is the Pearson's correlation coefficient:
\begin{equation}
	\rho(\tau) = \frac{\left\langle \left( I_1(t) - \left\langle I_1 \right\rangle \right) \cdot \left( I_2(t + \tau) - \left\langle I_2 \right\rangle \right) \right\rangle}{\sqrt{\langle \left( I_1(t) - \langle I_1 \rangle \right)^2 \rangle} \cdot \sqrt{\langle \left( I_2(t) - \langle I_2 \rangle \right) ^2 \rangle}}
	\label{eqn:pearson}
\end{equation}
It must be noted that the calculation of Pearson's correlation coefficient is non-trivial: MAGIC makes use of the convolution theorem for discrete Fourier transforms because it is computationally more efficient. Since $I_1$ and $I_2$ are proportional to the direct current of the PMTs ($DC_i$), the normalized contrast can be calculated as 
\begin{equation}
	c \propto \frac{\rho}{\sqrt{\left(DC_1 \cdot DC_2 \right)}}.
	\label{eq:norm_contrast2}
\end{equation}
One must also correct for a non-constant gain of the PMTs, as well as the moon. The latter can be done by measuring the background light with an additional PMT with no mirrors focused on it \citep{acciari2020optical}.

The significance of the signal can be expressed as the signal-to-noise ratio (SNR), which depends on many factors. However, most importantly, it is proportional to the absolute value of visibility, which itself depends on the distance between the telescopes \citep{acciari2020optical} 
\begin{equation}
	SNR = A \cdot \alpha \cdot q \cdot n \cdot \abs{V_{12}}^{2}(d) \cdot \sqrt{b_v} \cdot F^{-1} \cdot \sqrt{\frac{T}{2}} \cdot \sigma.
	\label{eq:SNR}
\end{equation}
Here, $A$ is the total mirror area, $\alpha$ the quantum efficiency of the PMTs, $q$ the quantum efficiency of the optics, $n$ the differential photon flux from the source, and $b_v$ the cross-correlation bandwidth. The noise of the PMTs is accounted with $F$, $T$ denotes the observation time, and $\sigma$ is the normalized spectral distribution of the light (including filters) \citep{acciari2020optical}. While most of the parameters can be optimized with hardware, the only way to obtain a better SNR is to increase the observation time with fixed telescopes.  
\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig/baseline.png}
	\caption{The tracking of baselines with four telescopes arranged in fig.~\ref{fig:teles} for one night of observation.}
	\label{fig:base}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{fig/ellipse/ellipse6018.jpg}
	\caption{This figure shows the simulated fast rotating star. The brightness is highest at the pole and there is gravitational darkening at the equator.}
	\label{fig:image}
\end{figure}
\subsection{Baseline considerations}
The measurement of the size of stellar objects through absolute visibility depends on the distance between the telescopes, which is called the baseline $B$. 
\begin{equation}
	V_{12}(B) = \frac{c(B)}{c(0)}
	\label{eq:angular_size_meas}
\end{equation}
However, this work needs a good SNR value for high precision of measurement, so the large covered observational plane with telescopes is a necessity for II \citep{acciari2020optical, abeysekara2020demonstration}. If the source is at the zenith, the coordinates in the Fourier plane ($u,v$) are given by:
\begin{equation}
	(u,v) = \frac{1}{\lambda} (B_N, B_E)
\end{equation}
where $B_N$ and $B_E$ are the baselines expressed in north and east coordinates. Since not all sources are at the zenith, and the telescopes are stationary, Earth's rotation plays an important role in covering the maximum observational plane through the rotated baselines. It is given as  
\begin{equation}
	\begin{pmatrix} u\\ v\\ w\\ \end{pmatrix} = R_x(\delta) \cdot R_y(h) \cdot R_x(-l) \begin{pmatrix} B_N\\ B_E\\ B_A\\ \end{pmatrix}
	\label{eq:baseline_rot}
\end{equation}
It traces an ellipse for every pair of telescopes. Furthermore, the different altitudes $B_A$ of the telescopes must be considered \citep{dravins2013optical, saha2020theory}. Here, $\delta$ is the declination and $h$ is the hour angle of the stellar source, and $l$ is the latitude of the telescopes. The three matrices $R_i$ correspond to the fundamental representation of the SO(3) group \citep{saha2020theory}. Fig.~\ref{fig:base} shows the track of six baselines generated from the telescopes (fig.~\ref{fig:teles}) due to Earth's rotation. Since every pair of telescopes traces an ellipse in the Fourier plane, the total number of ellipses scales as follows:
\begin{equation}
	\label{eq:N_telescopes}
	\mathcal{N} = \frac{N_T \cdot (N_T -1)}{2}
\end{equation}

As the number of baselines increases non-linearly, II benefits greatly from a large number of telescopes. So, the existing and upcoming Cherenkov Telescope Array Observatory (CTAO) will cover the maximum observational plane and provide insight into stellar objects with optical wavelengths in the future.
\begin{figure}
	\centering
	\begin{subfigure}{\linewidth}
		\includegraphics[width=\linewidth]{fig/ft/ft.jpg}
		\caption{The fourier transform of source.}
	\end{subfigure}\hfill
	\begin{subfigure}{\linewidth}
		\includegraphics[width=\linewidth]{fig/ft/ft_log.jpg}
		\caption{The logarithmic fourier transform of source.}
	\end{subfigure}
	\caption{Absolute value of the two-dimensional Fast Fourier Transform of fig.~\ref{fig:image} on linear and logarithmic scales. Observation of maximum (u, v) plane with finite number of baselines can be completed with help of earth's rotation.}
	\label{fig:ft}
\end{figure}
\begin{figure}
	\centering
	\begin{subfigure}{\linewidth}
		\includegraphics[width=\linewidth]{fig/ft/ft_base.jpg}
		\caption{The fourier transform with baselines.}
	\end{subfigure}\hfill
	\begin{subfigure}{\linewidth}
		\includegraphics[width=\linewidth]{fig/ft/ft_log_base.jpg}
		\caption{The logarithmic fourier transform with baselines.}
	\end{subfigure}
	\caption{The above panel shows the absolute value of the two-dimensional Fast Fourier Transform of fig.~\ref{fig:image} measured by baselines shown in fig.~\ref{fig:teles} on linear scale and below one shows the same on logarithmic scale.}
	\label{fig:ft_base}
\end{figure}
\subsection{An Object: Fast Rotating Stars}
In our work, we simulate a single fast-rotating star to test image reconstruction with II in intersection with the Generative Adversarial Networks (GAN). Fast rotation causes stars to take on an oblate shape, flattening at the poles and bulging at the equator due to the existence of stronger centrifugal force \citep{von1924radiative, 1999A&A...347..185M}. Fig.~\ref{fig:image} shows one of the simulations of a fast-rotating star, where the star takes on an elliptical shape with brightness distributed across the object. The brightness is maximum at the poles and minimum at the equator, a phenomenon known as gravity darkening \citep{lucy1967gravity}. This effect was first observed in the fast-rotating star Regulus through interferometric and spectroscopic data from the Center for High Angular Resolution Astronomy (CHARA) Array \citep{mcalister2005first}. These stars are crucial for understanding a variety of astrophysical processes and help to understand the stellar evolution, internal structure, and rotational dynamics over time.

Intensity Interferometry (II) detects the photon from the stellar objects, and the correlation of it results in squared visibility (explained in subsection.\ref{sec:signal}). According to Van Cittert Zernike's theorem, this signal is the Fourier transform of brightness distribution in the sky. Fig.~\ref {fig:ft} visualize this statement for the source shown in fig.~\ref{fig:image} using II on linear and logarithmic scales. However, current observational techniques are unable to capture the full signal, and only partial data can be collected with a single night's simulation, which has been shown in fig.~\ref{fig:ft_base}. It is the covered observational plane using the baselines shown in fig.~\ref{fig:teles} on linear and logarithmic scales, respectively. However, the upcoming CTAO promises to greatly expand the observable portion of the visibility plane for stellar objects, offering new insights into the mysteries of the universe.

>>>>

\section{Intensity Interferometry with Imaging Cherenkov Telescope Array}
Recently, Intensity Interferometry (II) has emerged again as an additional tool to resolve stellar objects as it solves some of the key limitations of classical optical interferometry (atmospheric turbulence, length of baseline, and the sensitivity of the optical aperture). It is suitable for the high-resolution study of very massive, hot, and short-lived stars, such as OB-type, Wolf-Rayet, and Pulsating stars in N-body systems. R. Hanbury Brown and Robert Q. Twiss first installed the Narrabri Stellar Intensity Interferometer (NSII) in 1974 to resolve the diameter of bright stars at visible wavelengths \citep{brown1954lxxiv, brown1974intensity}. Nowadays, Imaging Cherenkov Telescope Arrays (ICTAs), which are primarily used for gamma-ray observations, are at the forefront of implementing II during moonlit nights when it does not detect gamma-ray radiation. As a result, detection of intensity correlation and angular size measurements of stars is going on for highly resolved imaging with optical wavelength, particularly using Very Energetic Radiation Imaging Telescope Array Systems (VERITAS), High Energy Stereoscopic Systems (HESS), and Major Atmospheric Gamma Imaging Cherenkov Telescopes (MAGIC) \citep{kieda2022performance, zmija2022optical, abe2024performance}.

@@

\subsection{Signal-to-Noise Ratio for Intensity Interferometry}
\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig/telescope.png}
	\caption{Configuration of the four Imaging Cherenkov Telescope Array (ICTAs) with similar properties each used to simulate the signal for II.}
	\label{fig:teles}
\end{figure}
The main physics channel of ICTAs is the study of Very High Energy gamma rays ($\geq 30$ GeV) originating from particle showers in the atmosphere. Telescopes have an array of mirrors, which focus the light onto an array of photo-multiplier tubes (PMTs) \citep{aleksic2016major}. Here, Fig.~\ref{fig:teles} shows an arrangement of four ICTAs with similar properties each. For the simulation of II, the light of a stellar source is focused on a single PMT, which has a filter in front. The purpose of this filter is to efficiently transmit light around mean observational frequency, which protects the PMTs from excessive light. It is necessary, as II observations are mainly done during full moon nights when it is too bright for Very High Energy $\gamma$-ray observations. The optical signal of the PMTs is transmitted with optical fibers, converted to an electrical signal, and digitized \citep{acciari2020optical}. The measurable observable is the Pearson's correlation coefficient:
\begin{equation}
	\rho(\tau) = \frac{\left\langle \left( I_1(t) - \left\langle I_1 \right\rangle \right) \cdot \left( I_2(t + \tau) - \left\langle I_2 \right\rangle \right) \right\rangle}{\sqrt{\langle \left( I_1(t) - \langle I_1 \rangle \right)^2 \rangle} \cdot \sqrt{\langle \left( I_2(t) - \langle I_2 \rangle \right) ^2 \rangle}}
	\label{eqn:pearson}
\end{equation}
It must be noted that the calculation of Pearson's correlation coefficient is non-trivial: MAGIC makes use of the convolution theorem for discrete Fourier transforms because it is computationally more efficient. Since $I_1$ and $I_2$ are proportional to the direct current of the PMTs ($DC_i$), the normalized contrast can be calculated as 
\begin{equation}
	c \propto \frac{\rho}{\sqrt{\left(DC_1 \cdot DC_2 \right)}}.
	\label{eq:norm_contrast2}
\end{equation}
One must also correct for a non-constant gain of the PMTs, as well as the moon. The latter can be done by measuring the background light with an additional PMT with no mirrors focused on it \citep{acciari2020optical}.

The significance of the signal can be expressed as the signal-to-noise ratio (SNR), which depends on many factors. However, most importantly, it is proportional to the absolute value of visibility, which itself depends on the distance between the telescopes \citep{acciari2020optical} 
\begin{equation}
	SNR = A \cdot \alpha \cdot q \cdot n \cdot \abs{V_{12}}^{2}(d) \cdot \sqrt{b_v} \cdot F^{-1} \cdot \sqrt{\frac{T}{2}} \cdot \sigma.
	\label{eq:SNR}
\end{equation}
Here, $A$ is the total mirror area, $\alpha$ the quantum efficiency of the PMTs, $q$ the quantum efficiency of the optics, $n$ the differential photon flux from the source, and $b_v$ the cross-correlation bandwidth. The noise of the PMTs is accounted with $F$, $T$ denotes the observation time, and $\sigma$ is the normalized spectral distribution of the light (including filters) \citep{acciari2020optical}. While most of the parameters can be optimized with hardware, the only way to obtain a better SNR is to increase the observation time with a fixed number of telescopes.  
\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig/baseline.png}
	\caption{The tracking of baselines with four telescopes arranged in fig.~\ref{fig:teles} for one night of observation.}
	\label{fig:base}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{fig/ellipse/ellipse6018.jpg}
	\caption{This figure shows the simulated fast-rotating star. The brightness is highest at the pole, and gravitational darkening is present at the equator.}
	\label{fig:image}
\end{figure}
\subsection{Baseline considerations}
The measurement of the size of stellar objects through absolute visibility depends on the distance between the telescopes, which is called the baseline $B$. 
\begin{equation}
	V_{12}(B) = \frac{c(B)}{c(0)}
	\label{eq:angular_size_meas}
\end{equation}
However, this work needs a good SNR value for high precision of measurement, so the large covered observational plane with telescopes is a necessity for II \citep{acciari2020optical, abeysekara2020demonstration}. If the source is at the zenith, the coordinates in the Fourier plane ($u,v$) are given by:
\begin{equation}
	(u,v) = \frac{1}{\lambda} (B_N, B_E)
\end{equation}
where $B_N$ and $B_E$ are the baselines expressed in north and east coordinates. Since not all sources are at the zenith, and the telescopes are stationary, Earth's rotation plays an important role in covering the maximum observational plane through the rotated baselines, which is given as  
\begin{equation}
	\begin{pmatrix} u\\ v\\ w\\ \end{pmatrix} = R_x(\delta) \cdot R_y(h) \cdot R_x(-l) \begin{pmatrix} B_N\\ B_E\\ B_A\\ \end{pmatrix}
	\label{eq:baseline_rot}
\end{equation}
It traces an ellipse for every pair of telescopes. Furthermore, the different altitudes of the telescopes $B_A$ must be considered \citep{dravins2013optical, saha2020theory}. Here, $\delta$ is the declination, and $h$ is the hour angle of the stellar source, and $l$ is the latitude of the telescopes. The three matrices $R_i$ correspond to the fundamental representation of the SO(3) group \citep{saha2020theory}. Fig.~\ref{fig:base} shows the track of six baselines generated from the telescopes (fig.~\ref{fig:teles}) due to Earth's rotation. Since every pair of telescopes traces an ellipse in the Fourier plane, the total number of ellipses scales as follows:
\begin{equation}
	\label{eq:N_telescopes}
	\mathcal{N} = \frac{N_T \cdot (N_T -1)}{2}
\end{equation}

As the number of baselines increases non-linearly, II benefits greatly from a large number of telescopes. The existing and upcoming Cherenkov Telescope Array Observatory (CTAO) can promise to cover the maximum observational plane and provide insight into stellar objects with optical wavelengths in the future.
\begin{figure}
	\centering
	\begin{subfigure}{\linewidth}
		\includegraphics[width=\linewidth]{fig/ft/ft.jpg}
		\caption{The Fourier transform of the source.}
	\end{subfigure}\hfill
	\begin{subfigure}{\linewidth}
		\includegraphics[width=\linewidth]{fig/ft/ft_log.jpg}
		\caption{The logarithmic Fourier transform of the source.}
	\end{subfigure}
	\caption{Absolute value of the two-dimensional Fast Fourier Transform of Fig.~\ref{fig:image} on linear and logarithmic scales. Observation of the maximum (u, v) plane with a finite number of baselines can be completed with the help of Earth's rotation.}
	\label{fig:ft}
\end{figure}
\begin{figure}
	\centering
	\begin{subfigure}{\linewidth}
		\includegraphics[width=\linewidth]{fig/ft/ft_base.jpg}
		\caption{The Fourier transform with baselines.}
	\end{subfigure}\hfill
	\begin{subfigure}{\linewidth}
		\includegraphics[width=\linewidth]{fig/ft/ft_log_base.jpg}
		\caption{The logarithmic Fourier transform with baselines.}
	\end{subfigure}
	\caption{The upper panel shows the absolute value of the two-dimensional Fast Fourier Transform of Fig.~\ref{fig:image} measured by baselines shown in Fig.~\ref{fig:teles} on a linear scale, and the lower panel shows the same on a logarithmic scale.}
	\label{fig:ft_base}
\end{figure}
\subsection{An Object: Fast Rotating Stars}
In our work, we simulate a single fast-rotating star to test image reconstruction with II in intersection with the Generative Adversarial Networks (GAN). Fast rotation causes stars to take on an oblate shape, flattening at the poles and bulging at the equator due to the existence of stronger centrifugal force \citep{von1924radiative, 1999A&A...347..185M}. Fig.~\ref{fig:image} shows one of the simulations of a fast-rotating star, where the star takes on an elliptical shape with brightness distributed across the object. The brightness is maximum at the poles and minimum at the equator, a phenomenon known as gravity darkening \citep{lucy1967gravity}. This effect was first observed in the fast-rotating star Regulus through interferometric and spectroscopic data from the Center for High Angular Resolution Astronomy (CHARA) Array \citep{mcalister2005first}. These stars are crucial for understanding a variety of astrophysical processes and help to understand the stellar evolution, internal structure, and rotational dynamics over time.

Intensity Interferometry (II) detects the photon from the stellar objects, and the correlation of it results in squared visibility (explained in subsection.\ref{sec:signal}). According to Van Cittert Zernike's theorem, this signal is the Fourier transform of brightness distribution in the sky. Fig.~\ref {fig:ft} visualize this statement for the source shown in Fig.~\ref{fig:image} using II on linear and logarithmic scales. However, current observational techniques are unable to capture the full signal, and only partial data can be collected with a single night's simulation, which has been shown in Fig.~\ref{fig:ft_base}. It is the covered observational plane using the baselines shown in Fig.~\ref{fig:teles} on linear and logarithmic scales, respectively. However, the upcoming CTAO can promise to greatly expand the observable portion of the visibility plane for stellar objects, offering new insights into the mysteries of the universe.
